---
title: "Homework (FinKont)"
author: "Joakim Bilyk"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{wrapfig}
   - \usepackage{graphics}
output:
  html_document:
    toc: no
    code_folding: hide
  pdf_document:
    toc: yes
---

```{r, setup,include=FALSE}
library(knitr)
if (knitr::is_latex_output()) {
  knit_hooks$set(wrapf = function(before, options, envir) {
  if (!before) {
    output <- vector(mode = "character", length = options$fig.num + 1)

    for (i in 1:options$fig.num) {
      output[i] <- sprintf("\\includegraphics{%s}", fig_path(number = i))
    }

    output[i+1] <- "\\end{wrapfigure}"
    output <- c("\\begin{wrapfigure}{R}{0.5\\textwidth}",output)
    if (length(output)>= 3) {
      return(paste(output, collapse = ""))
    }
  }
})
  knitr::opts_chunk$set(
    fig.show = 'hide',
    echo = FALSE
  )
} else {
  knitr::knit_hooks$set(wrapf = function(before, options, envir) {})
  knitr::opts_chunk$set(
    out.extra = 'style="float:right; padding:10px"',
    out.width= "40%"
  )
}
```

```{r, include=FALSE, results = 'hide'}
library(ggplot2)
library(dplyr)
#rmarkdown::render(input = "FinKont_homework.Rmd", output_format = "pdf_document")
```


# Introduction

# Weeks {.tabset}

## Week 1

### Material

  * Brownian motion (Chapter 4.1)
  * Conditional expectation (Appendix B.5) 
  * Filtration (Appendix B.3 and Chapter 4.2)
  * Martingales (Appendix C.1 and Chapter 4.4)
  * Introduction (Chapter 1)
  * Discrete time models (Chapter 2 and 3)

### Theory

This week revolves around the theory of the Brownian motion and martingale processes. Other main topics are the binomial model and an introduction to financial derivatives. Financial derivatives is contingent on the outcome of a stochastic process at some future time $t=T$ and often is a function $\Phi$ of some assets price $S_t$. As such the derivative will give a stochastic payout, at time $t=T$ of the size $X_T=\Phi(S_T)$. Naturally we want to say something about the *fair* price of the derivative in the form of

$$\Pi_t(X_T)=\mathbb{E}\left[\Phi(S_T)\ \vert\ \mathcal{F}_t\right],$$

where $\mathcal{F}_t\subset\mathcal{F}$ is the available information at time $t$. We will by defualt intepret the times $t=0$ as *today* and $t=T$ as *tomorrow*. This indeed require some fundamental understanding of the behaviour of the asset price $S_t$. This lead us over to discussing the process in center of the *Black-Scholes* model: the Brownian motion.

&nbsp;

#### The Brownian motion

<blockquote class = "def">
**Definition 4.1.** *(Brownian motion)* A stochastic process $W$ is called a **Brownian motion** or **Wiener process** if the following conditions hold

 1. $W_0=0$.
 2. The process $W$ has independent increments, i.e. if $r<s\le t< u$ then $W_u-W_t$ and $W_s-W_r$ are independent random variables.
 3. For $s<t$ the random variable $W_t-W_s$ has the Gaussian distribution $\mathcal{N}(0,t-s)$.
 4. $W$ has continuous trajectories i.e. $s\mapsto W(s;\omega)$ i continuous for all $\omega \in\Omega$.
</blockquote>

```{r}
#Example of trajectory for BM
set.seed(1)
t <- 0:1000
N <- rnorm(
  n = length(t)-1, #initial value = 0
  mean = 0, #incements mean = 0
  sd = sqrt(t[2:length(t)] - t[1:(length(t)-1)]) #increment sd = sqrt(t-s)
)
W <- c(0,cumsum(N))
```
```{r,echo=FALSE,include=TRUE,wrapf = TRUE}
data.frame(t = t, W = W) %>%
  ggplot() +
  geom_line(aes(x=t,y=W)) +
  labs(title = "Realisation of a Brownian motion") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        title = element_text(size = 18))
```

As one can see from the simulated sample path on the right, the Brownian motion is rather irratic. In fact, the process varies infinitely on any interval with length greater than 0. This gives some of the characteristics of the process including that: $W$ is continuous and $W$ is non-differential everywhere. This irratic behaviour is summed up in the theorem.


<blockquote class = "thm">

**Theorem 4.2.** A Brownian motions trajectory $t\mapsto W_t$ is with probability one nowhere differential, and it has locally infinite total variation.

</blockquote>

This may seem not that horrifying since we can observe the process at any time and conclude an increment $W_{t+\Delta t}-W_t$ for any $\Delta t>0$ but any integral constructed with $W_t$ as integrator becomes nonsensible. We will be studying processes on the form

$$S_{t+\Delta t}-S_t=\mu(t,S_t)\Delta t+\sigma(t,S_t)\Delta W_t,\hspace{20pt}\Delta W_t=W_{t+\Delta t}-W_t.$$

where $W_t$ is a standard Brownian motion and $\mu(t,S_t)$ is locally deterministic (velocity), that is $\mu(t,S_t)$ is deterministic on a small time interval. One could consider the dynamics of the process $S_t$ by studying the equation below as $\Delta t\to 0_+$

$$\frac{S_{t+\Delta t}-S_t}{\Delta t}=\mu(t,S_t)+\sigma(t,S_t)\frac{W_{t+\Delta t}-W_t}{\Delta t}.$$

The limit however is impossible to determine as $W_t$ is non-differentiable and as such $dS_t$ is not well-defined. From LivStok we know that the dynamics of $S_t$ is given by letting $\Delta t$ tend to 0 without dividing by it, that is

$$
dS_t=\mu(t,S_t)\ dt+\sigma(t,S_t)\ dW_t.
$$

Giving that $S_0$ is observable we could intepret the dynamics on the integral form

$$
S_t=S_0+\int_0^t\mu(s,S_s)\ ds+\int_0^t\sigma(s,S_s)\ dW_s,
$$

where the above integrals is Riemann-Stieltjes integral. This is however still at dead-end, since from theorem 4.2 we know that $W_t$ has unbounded variation on any interval. So **we cannot define $S_t$ for each $W$-trajectory seperately** we will despite this define another integral (the Ito integral) that in some other sense give a global solution to this integral. To this we will be considering a $L^2$-definition.

&nbsp;

#### Conditional expectation

The theory of conditional expectation is well-known from courses on the bachelor. Because of this we will only summarise the most important results.

We consider a background space $(\Omega,\mathcal{F},P)$ and a sub-sigma algebra $\mathcal{G}\subseteq \mathcal{F}$. We assume that some stochastic variable is $\mathcal{F}$-measurable, that is the mapping $X : (\Omega,\mathcal{F},P) \to (\mathbb{R},\mathbb{B},m)$ is $\mathcal{F}-\mathbb{B}$-measurable i.e. $\forall B\in\mathbb{B} : \{X\in B\}\in\mathcal{F}$. For some random variable $Z$ defined on the subspace $(\Omega,\mathcal{G},P)$, we say that $Z$ is the conditional expectation of $X$ given $\mathcal{G}$ if

$$
\forall G\in\mathcal{G} : \int_G Z(\omega)\ dP(\omega)=\int_G X(\omega)\ dP(\omega).
$$

This fact is summed up in the definition below.

<blockquote class = "def">

**Definition B.27.** *(Conditional expectation)* Let $(\Omega,\mathcal{F},P)$ be a probability space and $X$ a random variable in $L^1(\Omega,\mathcal{F},P)$ ($\vert X\vert$ is integrable). Let furthermore $\mathcal{G}$ be a sigma-algebra such that $\mathcal{G}\subseteq \mathcal{F}$. If $Z$ is a random variable with the properties that:

  i. $Z$ is $\mathcal{G}$-measurable.
  ii. For every $G\in\mathcal{G}$ it holds that
  $$\int_G Z(\omega)\ dP(\omega)=\int_G X(\omega)\ dP(\omega).$$

Then we say that $Z$ is the *conditional expectation of $X$ given the sigma-algebra $\mathcal{G}$*. In that case we denote $Z$ by the symbol $E[X\ \vert\ \mathcal{G}]$.

</blockquote>

We see that from the above it always holds that $X$ satisfies (ii). It does not, however, always hold that $X$ is $\mathcal{G}$-measurable. Given this fact it is not trivial that a random variable $E[X\ \vert\ \mathcal{G}]$ even exists. This nontriviality is fortunatly resolved by the Radon-Nikodym theorem.

<blockquote class = "thm">

**Theorem B.28.** *(Existance and uniqueness of Conditional expectation)* Let $(\Omega,\mathcal{F},P)$, $X$ and $\mathcal{G}$ be given as in the definition above. Then the following holds:

  * There will always exist a random variable $Z$ satisfying conditions (i)-(ii) above.
  * The variable $Z$ is unique, i.e. if both $Y$ and $Z$ satisfy (i)-(ii) then $Y=Z$ $P$-a.s.

</blockquote>

This result ensures that we may condition on any sigma-algebra for instance $\mathcal{G}=\sigma(Y)$ in that case we (pure notation) write

$$
E[X\ \vert\ \sigma(Y)]=E[X\ \vert\ Y],\hspace{20pt}\sigma(Y)=\sigma\left(\left\{ Y\in A,\ A\in\mathbb{B}\right\}\right).
$$

In the above $\sigma(Y)$ is simply the smallest sigma-algebra containing all the pre-images of $Y$, that is the smallest sigma-algebra making $Y$ measurable! Giving this foundation there are a few properties conditional expectation have which is rather useful (for instance the tower property).

Below we assume: Let $(\Omega,\mathcal{F},P)$ be a probability space and $X,Y$ be random variables in $L^1(\Omega,\mathcal{F},P)$.

<blockquote class = "prop">

**Proposition B.29.** *(Monotinicity/Linearity of Conditional expectation)* The following holds:

$$X\le Y\ \Rightarrow\ E[X\ \vert\ \mathcal{G}]\le E[Y\ \vert\ \mathcal{G}],\hspace{20pt}P-\text{a.s.}$$
$$E[\alpha X + \beta Y\ \vert\ \mathcal{G}]=\alpha E[X\ \vert\ \mathcal{G}]+ \beta E[Y\ \vert\ \mathcal{G}],\hspace{20pt}\forall \alpha,\beta\in\mathbb{R}.$$

</blockquote>

<blockquote class = "prop">

**Proposition B.30.** *(Tower property)* Assume that it holds that $\mathcal{H}\subseteq\mathcal{G}\subseteq\mathcal{F}$. Then the following hold:

$$E[E[X\vert \mathcal{G}]\vert\mathcal{H}]=E[X\vert \mathcal{H}],$$
$$E[X]=E[E[X\vert \mathcal{G}]].$$

</blockquote>

<blockquote class = "prop">

**Proposition B.31.** Assume $X$ is $\mathcal{G}$ and that both $X,Y$ and $XY$ are in $L^1$ (only assuming $Y$ is $\mathcal{F}$-measurable), then

$$E[X\vert\mathcal{G}]=X,\hspace{20pt}P-\text{a.s.}$$
$$E[XY\vert\mathcal{G}]=XE[Y\vert\mathcal{G}],\hspace{20pt}P-\text{a.s.}$$

</blockquote>

<blockquote class = "prop">

**Proposition B.32.** *(Jensen inequality)* Let $f:\mathbb{R}\to\mathbb{R}$ be a convex (measurable) function and assume $f(X)$ is in $L^1$. Then

$$f(E[X\vert\mathcal{G}])\le E[f(X)\vert\mathcal{G}],\hspace{20pt}P-\text{a.s.}$$

</blockquote>

&nbsp;

#### Filtrations

Let $(\Omega,\mathcal{F},P)$ be a probability space. We define a filtration as an increasing famility of sub-sigma-algebras in the following definition.

<blockquote class = "def">

**Definition B.16.** *(Filtration)* Let $\mathbf{F}=(\mathcal{F}_t)_{t\ge 0}$ be an time indexed family of sub-sigma-algebras such that $F_s\subseteq F_t$ for $s\le t$ and $\mathcal{F}_t\subseteq \mathcal{F}$ for all $t\ge 0$. We may given this filtration define $\mathcal{F}_\infty$ as $\sigma\left(\bigcup_{t\ge 0}\mathcal{F}_t\right)$.

</blockquote>

Filtrations is widely used in stochastic processes, as they allow for the concept of knowledge/information. This is useful when considering mean-values of future states but in an increasing information setting. For this we introduce the term adapted processes.

<blockquote class = "def">

**Definition B.17.** *(Adapted process)* Let $(\mathcal{F}_t)_{t\ge 0}$ be a filtration on the probability space $(\mathcal{F}_t)_{t\ge 0}$. Furthermore, let $(X_t)_{t\ge 0}$ be a stochastic process on the same space. We say that $X_t$ is adapted to the filtration $\mathbf{F}$ if

$$X_t\ \text{ is }\ \mathcal{F}_t-\text{measurable},\hspace{20pt}\forall t\ge 0.$$

</blockquote>

Obviously, we may introduce the **natural filtration** $\mathcal{F}^X_t$ given by the tragetory of the process $X_t$:

$$\mathcal{F}^X_t=\sigma(\{X_s,\ s\le t\}).$$

Indeed, $X_t$ is adapted to this filtration.

&nbsp;

#### Martingales

<blockquote class = "def">

**Definition C.1.** Let $M_t$ be a stochastic process defined on a background space $(\Omega,\mathcal{F},P)$. Let $(\mathcal{F}_t)_{t\ge 0}$ be a filtration. If $M_t$ is adapted to the filtration $\mathcal{F}_t$, $E\vert M_t\vert <\infty$ and

$$E[M_t\vert \mathcal{F}_s]=M_s,\hspace{20pt}P-\text{a.s.}$$

holds for any $t>s$ we say that $M_t$ is a martingale ($\mathbf{F}$-martingale). If the above has $\le$ or $\ge$ we say that $M_t$ is either a **submartingale** or **supermartingale** respectively.

</blockquote>

Naturally, this defintions may easily be extended to discrete models and we have the trivial equality:

$$E[M_t-M_s\ \vert\ \mathcal{F}_s]=0.$$

Martingales is useful, when proofing probalistic statements as the posses tractable properties. A useful technique often include the construction of the martingale

$$M_t=E[X\ \vert\ \mathcal{F}_t].$$

&nbsp;

#### Discrete time models

#### One-period time models

The study of this course is the **European call** option (and *put* option). This financial derivative is an agreement between two parties where the holder of the option has the right to *"exercise"* the derivative, at a future time $t=T$. Exercising means buying an asset at a certain agreed opon price-strike $K$. In the case of the put-option: the holder has the right (but not obligation) to sell the asset at the strike price $K$. As such the derivative has the payoff

$$\text{Call}\ \text{option:}\hspace{10pt}\Phi(S_T)=(S_T-K)^+,\hspace{20pt}\text{Put}\ \text{option:}\hspace{10pt}\Phi(S_T)=(K-S_T)^+.$$

Our objective is to understand when an arbitrage exist and to find the fair price of these derivative. The strategy in pricing is finding a replicating portfolio with the same payoff as the option (with probability one) and then price the derivative accordingly.

##### Model description

In the one-period model we consider the simplest possible market. We have two distinct times $t=0$ (today) and $t=1$ (tomorrow) and we may buy any portfolio as a mixture of bonds and one stock. We denote the bonds price by $B_t$ and the stocks price by $S_t$ and we assume the following:

$$
B_0=1,\ B_1=1+R,\hspace{20pt}S_0=s,\ S_1=\left\{\begin{matrix}s\cdot u, & with\ probability\ p_u.\\s\cdot d, & with\ probability\ p_d.\end{matrix}\right.
$$

We may introduce $Z$ as the random variable

$$
Z=u\cdot (I)+d\cdot (1-I),
$$

for an bernoulli variable $I$ with succes probability $p_u$. Naturally, we assume $d\le (1+R)\le u$ (this is imperative to ensure no arbitrage as we will see).

##### Portfolios and arbirtage

We study any portfolio on the $(B,S)$ market as a vector $h=(x,y)$ where $x$ is the amount of bonds and $y$ is the amount of stock held in the portfolio. Notice that we allow for shorting, that is $x<0$ or $y<0$. As such, we have that $h\in \mathbb{R}^2$. In this we have made some unrealistic, but attractable assumptions included in the assumptions:

  * We allow short positions and fractional holding, i.e. $h\in \mathbb{R}^2$,
  * We assume no spread between ask and bids,
  * No transaction costs and
  * A completely liquid market i.e. we may borrow and buy as much stock and bonds as wanted.

Given that we have chosen a portfolio $h$ we may introduce the value process.

<blockquote class = "def">

**Definition 2.1.** The **value process** of the porfolio $h\in\mathbb{R}^2$ is the stochastic process

$$V^h_t=xB_t+yS_t,\ t=0,1.$$

</blockquote>

Given this notation we may define what an arbitrage is.

<blockquote class = "def">

**Definition 2.2.** An **arbitrage** is a portfolio $h$ with the properties: 1) $V^h_0=0$, 2) $P(V^h_1\ge 0)=1$ and 3) $P(V^h_1>0)>0$.

</blockquote>

That is $h$ is an deterministic money-machine where we at least never loose any money. Granted the bonds give a determinictic non-negative return, but an arbitrage does not require any money out of pocket. With the notion of an arbitrage we will show the first proposition regarding the choice of $R,u,d$ as defined above.

<blockquote class = "prop">

**Proposition 2.3.** The one-period binomial model is arbitrage free if and only if the following inequality hold:

$$d\le (1+R)\le u.\tag{2.1}$$

</blockquote>

<details>
<summary>**Proof.**</summary>

The statement is proofed by contradiction. Assume that $d>1+R$ holds. Then by definition $u>d>1+R$. Notice that any portfolio satisfying $V_0^h=0$ must satisfy

$$0=xB_0+yS_0=x+ys\iff x=-ys$$

That is for some choice $y$ the only arbitrage candidate is the portfolio $h=(-ys,y)$. Calculating the value at time $t=1$ we have

$$V_1^h=-ys\cdot(1+R)+y\cdot s\cdot Z=ys(Z-1-R)$$

However since $Z\ge d$ we have $Z-(1+R)\ge 0$ and therefore an arbitrage (for $y>0$). The other inequality $1+R>u$ follows analog steps. Simply choose some $y<0$ and the result follows. $\blacksquare$

</details>

From inequality (2.1) we see that since $1+R$ is between $u$ and $d$ we may find a pair $q_d,q_u\ge 0$ with $q_d+q_u=1$ such that

$$1+R=q_u\cdot u+q_d\cdot d.$$

This yields the important risk neutral valuation formula as summed op in the following definition

<blockquote class = "def">

**Definition 2.4.** A probability measure $Q$ is called a **martingale meausre** if the following condition holds:

$$S_0=\frac{1}{1+R}E^Q[S_1].$$

</blockquote>

The above measure $Q$ is the measure $Q(Z=d)=q_d$ and $Q(Z=u)=q_u$ for the binomial model. This does in fact yield the risk neautral valuation formula:

\begin{align*}
S_0&=\frac{1}{1+R}E^Q[S_1]=\frac{1}{1+R}(Q(Z=d)\cdot d\cdot s+Q(Z=u)\cdot u\cdot s)\\
&=s\frac{1}{1+R}(q_d\cdot d+q_u\cdot u)=s,
\end{align*}

where we simply use $1+R=q_d\cdot d+q_u\cdot u$. We call this the risk neautral valuation formula because it in some sense gives an expected discounted value of the future stock price. We end this endavour with reformulating the arbitrage proposition and determining the values of the $Q$-measure.

<blockquote class = "prop">

**Proposition 2.5.** The one-period binomial model is arbitrage free if and only if there exists a martingale measure $Q$.

</blockquote>

<blockquote class = "prop">

**Proposition 2.6.** The one-period binomial model has martingale probabilities given by:

$$\left\{\begin{matrix}q_u=\frac{(1+R)-d}{u-d},\\ q_u=\frac{u-(1+R)}{u-d}.\end{matrix}\right.$$

</blockquote>

##### Contingent Claims

This chapter revolves around the financial derivative and we start by stating the definition of the financial derivative.

<blockquote class = "def">

**Definition 2.7.** A **contingent claim** (financial derivative) is *any* stochastic variable $X$ of the form $\Phi(Z)$, where $Z$ is the stochastic varible driving the stock price process.

</blockquote>

We may also call the function $\Phi$ the **contract function** as it states how the contract is resolved once the stochastic variable $Z$ has been realised. Our objective is now to study, what a buyer of said contract would have to pay at any given time $t$. We call the fair price of $X$ at time $t$: $\Pi_t[X]$. As such it is easy to see that the fair price at the time of maturity $T$ is simply the payout $X$ i.e. $\Pi_T[X]=X$. Our strategy is to find a replicating portfolio $h$ and determine the price of said portfolio.

<blockquote class = "def">

**Definition 2.8.** A contingent claim $X$ can be **replicated**, or said to be **reachable** if there exist a portfolio $h$ such that

$$
V_1^h=X,
$$

with probability one. In that case, we say that the portfolio $h$ is a **hedging** portfolio or a **replicationg** portfolio. If all claims can be replicated we say that the market is **complete**.

</blockquote>

Our pricing strategy is then to determine the value process of the replicating portfolio and then by the first pricing principle below we say that the price is imply the value of the replicating portfolio.

**Pricing principle 1.** If a clain $X$ is reachable with replicating portfolio $h$, then the only reasonable price process for $X$ is given by

$$
\Pi_t[X]=V_t^h.
$$

Notice, that this assumes that a replicating portfolio exist and even so we have a uniqueness statement to solve. We end this section by writing two important results.

<blockquote class = "prop">

**Proposition 2.9.** Suppose that a claim $X$ is reachable with replicating portfolio $h$. Then any price at time $t\ge 0$ of the claim $X$ other than the value process of $h$ will lead to an arbitrage on the extended market $(B,S,X)$.

</blockquote>

<blockquote class = "prop">

**Proposition 2.10.** If the one-period binomial model is free of arbitrage, then it is also complete.

</blockquote>

The hedging portfolio in the one-period binomial model is given by the portfolio $(x,y)$ below

$$
x=\frac{1}{1+R}\cdot\frac{u\Phi(d)-d\Phi(u)}{u-d},\hspace{20pt}y=\frac{1}{s}\cdot\frac{\Phi(u)-\Phi(d)}{u-d}.
$$

##### Risk Neutral Valuation

We see that since the one-period model is complete we can price any contingent claim and we see that

\begin{align*}
\Pi_0[X]&=\frac{1}{1+R}\cdot\frac{u\Phi(d)-d\Phi(u)}{u-d}+s\frac{1}{s}\cdot\frac{\Phi(u)-\Phi(d)}{u-d}\\
&=\frac{1}{1+R}\left\{\frac{u\Phi(d)-d\Phi(u)}{u-d}+(1+R)\frac{\Phi(u)-\Phi(d)}{u-d}\right\}\\
&=\frac{1}{1+R}\left\{\frac{(1+R)-d}{u-d}\Phi(u)+\frac{u-(1+R)}{u-d}\Phi(d)\right\}\\
&=\frac{1}{1+R}E^Q[X].
\end{align*}

i.e. the price at time $t=0$ should simply be the expected discounted payout according to the martingale measure. This leads to the important pricing proposition:

<blockquote class = "prop">

**Proposition 2.11.** If the one-period binomial model is free of arbitrage, then the arbitrage free price of a contingent claim $X$ is given by

$$
\Pi_0[X]=\frac{1}{1+R}E^Q[X].\tag{2.4}
$$

Here the martingale measure $Q$ is uniquely determined by the relation

$$
S_0=\frac{1}{1+R}E^Q[S_1],\tag{2.5}
$$

and the explicit expressions for $q_u$ and $q_d$ are given in proposition 2.6. Furthermore the claim $X$ can be replicated using the portfolio

\begin{align*}
x&=\frac{1}{1+R}\cdot\frac{u\Phi(d)-d\Phi(u)}{u-d},\tag{2.6}\\
y&=\frac{1}{s}\cdot\frac{\Phi(u)-\Phi(d)}{u-d}.\tag{2.7}
\end{align*}

</blockquote>

#### Multi-period model

The one-period binomial model can easily be extended to a multi-period model, by assuming that the bond and stock pricess evolve by the processes:

$$
t\ge1:\ B_t=(1+R)B_{t-1}\hspace{20pt}\text{and}\hspace{20pt}B_0=1,
$$

$$
t\ge1:\ S_t=Z_{t-1}S_{t-1}\hspace{20pt}\text{and}\hspace{20pt}S_0=s,
$$

where we obviously have that $B_t=(1+R)^t$ for $t\ge 0$. In the above $Z_t$ is $u$ with probability $p_u$ and $d$ with probability $p_d$. In this context, we need to define a portfolio in terms of a strategy.

<blockquote class = "def">

**Definition 2.13.** A **portfolio strategy** is a stochastic process on $\{1,...,T\}$

$$
h=\left\{h_t=(x_t,y_t);\ t=1,...,T\right\}
$$

such that $h_t$ is a function of $S_0,S_1,...,S_{t-1}$. For a given portfolio strategy $h$ we set $h_0=h_1$ by convention. The associated **value process** corresponding to the portfolio $h$ is defined by

$$
V_t^h=x_t(1+R)+y_tS_t.
$$

</blockquote>

Given this notation we may define what an arbitrage is, but first we introduce the notion of a self-financing portfolio. A self-financing portfolio in an intuative sense is a portfolio that is not withdrawn from or deposited into.

<blockquote class = "def">

**Definition 2.14.** A portfolio strategy $h$ is said to be **self-financing** if the following condition holds for all $t=0,...,T-1$:

$$
x_t(1+R)+y_tS_t=x_{t+1}+y_{t+1}S_t.
$$

</blockquote>

The above equation says that the portfolio purchased at time $t$ and helt until $t+1$ $(x_{t+1},y_{t+1})$ can only be financed by the market value of the portfolio held from $[t-1,t)$ i.e. $(x_{t},y_{t})$. We now define an arbitrage.

<blockquote class = "def">

**Definition 2.15.** An **arbitrage** is a self-financing portfolio $h$ with the properties: 1) $V^h_0=0$, 2) $P(V^h_T\ge 0)=1$ and 3) $P(V^h_T>0)>0$.

</blockquote>

The multiperiod binomial model has an just like the oneperiod model a result regarding when an arbitrage exists.

<blockquote class = "lem">

**Lemma 2.16.** If $d\le (1+R)\le u$ then the multiperiod model is arbitrage-free.

</blockquote>

As one can see, the multiperiod model is rather similar to the one period model. We wil in the following summarise equivalent statements for the multiperiod model as the ones in the oneperiod model. 

<blockquote class = "def">

**Definition 2.17.** The martingale probabilities $q_u$ and $q_d$ are defined as the probabilities for which the relation below holds.

$$
s=\frac{1}{1+R}E^Q[S_{t+1}\ \vert\ S_t].
$$

</blockquote>

<blockquote class = "prop">

**Proposition 2.18.** The martingale probabilities $q_u$ and $q_d$ are given by

$$
\left\{\begin{matrix}q_u=\frac{(1+R)-d}{u-d},\\ q_u=\frac{u-(1+R)}{u-d}.\end{matrix}\right.
$$

</blockquote>

<blockquote class = "def">

**Definition 2.19.** A **contingent claim** is a stochastic variable $X$ of the form

$$
X=\Phi(S_T),
$$

where the **contract function** $\mathbf{\Phi}$ is some given real valued function.

</blockquote>

<blockquote class = "def">

**Definition 2.20.** A given contingent claim $X$ is said to be **reachable** if there exists a self-financing portfolio $h$ such that

$$
V_T^h=X,
$$

with probability one. In that case we say that the portfolio $h$ is a **hedging** portfolio or a **replicating** portfolio. If all claims can be replicated we say that the market is *(dynamically)* **complete**.

</blockquote>

**Pricing principle 2.** If a claim $X$ is reachable with replicating portfolio $h$, then the only reasonable price process for $X$ os given by

$$
\Pi_t[X]=V_t^h,\ t=0,1,...,T.
$$

<blockquote class = "prop">

**Proposition 2.21.** Assume $X$ is reachable by $h$, then any price other than $V_t^h$ for some $t\ge 0$ leads to an arbitrage opportunity.

</blockquote>

<blockquote class = "prop">

**Proposition 2.22.** The multiperiod model is complete, i.e. every claim can be replicated by a self-financing portfolio.

</blockquote>

<blockquote class = "prop">

**Proposition 2.24.** **(Binomial algorithm)** Consider a $T$-claim $X=\Phi(S_T)$. Then this claim can be replicated using af self-financing portfolio. If $V_t(k)$ denotes the value of the portfolio at the node $(t,k)$ ($k$ referring to $k$ amount of up-moves for the stock), then $V_t(k)$ can be computed recursively by the scheme

$$
\left\{\begin{matrix}V_t(k)=\frac{1}{1+R}\left\{q_uV_{t+1}(k+1)+q_dV_{t+1}(k)\right\},\\ V_T(k)=\Phi(su^kd^{T-k}).\end{matrix}\right.
$$

where the martingale probabilities $q_u$ and $q_d$ are given by

$$
\left\{\begin{matrix}q_u=\frac{(1+R)-d}{u-d},\\ q_u=\frac{u-(1+R)}{u-d}.\end{matrix}\right.
$$

With the notation as above, the hedging portfolio is given by

$$
\left\{\begin{matrix}x_t(k)=\frac{1}{1+R}\cdot\frac{uV_t(k)-dV_t(k+1)}{u-d},\\ y_t(k)=\frac{1}{S_{t-1}}\cdot\frac{V_t(k+1)-V_t(k)}{u-d}.\end{matrix}\right.
$$

In particular, the arbitrage free price of the claim at $t=0$ is given by $V_0(0)$.

</blockquote>

<details>
<summary>**Example.**</summary>

```{tikz,echo=FALSE}
\tikzstyle{level 1}=[level distance=4cm, sibling distance=3.5cm,->]
\tikzstyle{level 2}=[level distance=4cm, sibling distance=2cm,->]

\tikzstyle{bag} = [text width=2em, text centered]
\tikzstyle{end} = []

\begin{tikzpicture}[grow=right, sloped]
\node[bag] {100}
    child {
        node[bag] {90}        
            child {
                node[end, label=right:
                    {81}] {}
                edge from parent
                node[above] {}
                node[below]  {$p_d^2$}
            }
            child {
                node[end, label=right:
                    {99}] {}
                edge from parent
                node[above] {}
                node[below]  {$p_dp_u$}
            }
            edge from parent 
            node[above] {}
            node[below]  {$p_d$}
    }
    child {
        node[bag] {110}        
        child {
                node[end, label=right:
                    {99}] {}
                edge from parent
                node[above] {$p_dp_u$}
                node[below]  {}
            }
            child {
                node[end, label=right:
                    {121}] {}
                edge from parent
                node[above] {$p_u^2$}
                node[below]  {}
            }
        edge from parent         
            node[above] {$p_u$}
            node[below]  {}
    };
\end{tikzpicture}
```

Consider $R=0.04$, $s=100$, $u=1.1$, $d=0.9$, $p_u=0.6$ and $p_d=0.4$. We consider a model of length $T=2$ and we want to evaluate the price of the european call option with srike $K=90$ that is the contingent claim

$$
X=(S_T-K)^+,\hspace{20pt}\Phi(s)=(s-K)^+.
$$

For each time $t$ we know the replicating portfolio, if we know the payoff the following period. Therefore we start from the leaves of the tree and work towards the root. Since the strike price is $K=90$ the end result will be the following payoffs:

\begin{align*}
u^2:\hspace{20pt}&(121-90)^+=31\\
ud:\hspace{20pt}&(99-90)^+=9\\
du:\hspace{20pt}&(99-90)^+=9\\
d^2:\hspace{20pt}&(81-90)^+=0
\end{align*}

Therefore by the risk neautral valuation formula with $q_u=\frac{(1+R)-d}{u-d}=0.7$ and $q_d=\frac{u-(1+R)}{u-d}=0.3$ we have that the cost of the replicating portfolio at time $t=1$ is respectively

\begin{align*}
u:\hspace{20pt}&\frac{1}{1+R}\left\{31\cdot q_u + 9 \cdot q_d\right\}\approx 23.46\\
d:\hspace{20pt}&\frac{1}{1+R}\left\{9\cdot q_u + 0 \cdot q_d\right\}\approx 6.06
\end{align*}

To replicate this payoff at time $t=1$ we can use the risk neutral valuation formula once more to find the base cost of the replicating portfolio i.e. the price of $X$ at time $t=0$

$$
\frac{1}{1+R}\left\{23.46\cdot q_u + 6.06 \cdot q_d\right\}\approx 17.54.
$$

Working from the root to the leaves we can now calculate the hedging portfolio at time $t=0,1$ for each path. For time $t=0$ we calculate

\begin{align*}
x=&\frac{1}{1+R}\cdot \frac{u\cdot 6.06-d\cdot 23.46}{u-d}\approx -69.46,\\
y=&\frac{1}{s}\cdot\frac{23.46-6.06}{u-d}\approx0.87
\end{align*}

We see by calculations that this does indeed replicate the payoff at time $t=1$:

\begin{align*}
u:\hspace{20pt}&V_1^h=(1+R)\cdot x + 110\cdot y\approx 23.46,\\
d:\hspace{20pt}&V_1^h=(1+R)\cdot x + 90\cdot y\approx 6.06.
\end{align*}

We also see by calculation that the initial portfolio does cost the expected 17.54 as

$$
x\cdot 1+y\cdot100=87-69.46=17.54.
$$

Following these steps at time $t=1$ the portfolios $(-86.54,1)$ (for the up-scenario) and $(-38.94,0.5)$ (for the down-scenario) would arise. Notice when calculating $y$ one has to use the current price $S_1=S_0\cdot Z$ not $S_0$. One should also check by similar calculations as above, that these portfolios does indeed replicate the payoff of the contingent claim $X$. $\square$

</details>

<blockquote class = "prop">

**Proposition 2.25.** The arbitrage free price at $t=0$ of a $T$-claim $X$ is given by

$$
\Pi_0[X]=\frac{1}{(1+R)^T}E^Q[X]
$$

where $Q$ denotes the martingale measure, or more explicitly

$$
\Pi_0[X]=\frac{1}{(1+R)^T}\sum_{k=0}^T\binom{T}{k}q_u^kq_d^{T-k}\Phi(su^kd^{T-k}).
$$

</blockquote>

<details>
<summary>**Example.**</summary>

```{r}
R <- 0.04
s <- 100
u <- 1.1
d <- 0.9
p_u <- 0.6
p_d <- 0.4
q_u <- (1+R-d)/(u-d)
q_d <- (u-1-R)/(u-d)
cap_t <- 2

#Test for K=90
K <- 90
pi_0 <- (1+R)**(-cap_t)*sum(
  choose(cap_t,0:cap_t)*q_u**(0:cap_t)*q_d**(cap_t - 0:cap_t)*pmax(s*u**(0:cap_t)*d**(cap_t - 0:cap_t)-K,0)
) # = 17.53883

pi_0 <- unlist(lapply(0:ceiling(s*u**cap_t), function(K){
  (1+R)**(-cap_t)*sum(
    choose(cap_t,0:cap_t)*q_u**(0:cap_t)*q_d**(cap_t - 0:cap_t)*pmax(s*u**(0:cap_t)*d**(cap_t - 0:cap_t)-K,0)
  )
}))
```
```{r, echo = FALSE, wrapf = TRUE}
library(dplyr)
library(ggplot2)
data.frame(K = 0:ceiling(s*u**cap_t),
           Pi_0 = pi_0) %>%
  ggplot(.) + geom_line(aes(x=K,y=Pi_0)) +
  labs(title = "Price of european call option",
       x = "Strike", y = "Price") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        title = element_text(size = 18))
```

We follow an analog example as the one after proposition 2.24. Let $K=90$ and we see that

\begin{align*}
&\Pi_0[X]\\
&=\frac{1}{(1+0.04)^2}\sum_{k=0}^2\binom{2}{k}\cdot0.7^k\cdot0.3^{2-k}\cdot\Phi(100\cdot 1.1^k\cdot0.9^{2-k})\\
&=0.9245562\cdot\left(\underbrace{1\cdot 1\cdot0.09\cdot0}_{k=0}+\underbrace{2\cdot 0.7\cdot0.
3\cdot 9}_{k=1}+\underbrace{1\cdot 0.49\cdot1\cdot31}_{k=2}\right)\\
&=0.9245562\cdot\left(0+3.78+15.19\right)\\
&=17.53883
\end{align*}


Since we know that $K$ must meaningfully range in $[0,121]$ we could try to calculate the price of the contingent claim at time $t=0$ for all integers in this interval. We see that the price range between $S_0$ and 0 as expected. One can also see that the price changes slope at the prices 99 and 121 as the function is linear in $\Phi$ and som realisations loose any effect on the price when the strike is higher than the outcome. $\square$

</details>

<blockquote class = "prop">

**Proposition 2.26.** The condition $d<(1+R)<u$ is necessary and sufficient condition for absence of arbitrage.

</blockquote>

#### Generelised one-period model

In the previous we had the simpel model where we only had one stochastic asset $S$ and only one stochastic variable $Z$ determining the future stock price. Now we will generelise this model by introducing $N$ assets and introducing som stochastic behaviour to the system.

##### Model specification

We consider the market consisting of a collection of stochastic prices assets $i=1,...,N$ with $N$-dimensional price process.

$$
S_t=\begin{bmatrix} S_t^1\\
\vdots\\
S_t^N\end{bmatrix}
$$

We now assume that $S_t$ is defined on a background space with finite sample space $\Omega = \{\omega_1,...,\omega_M\}$ with associated probabilities $p_j=P(\omega_j)$, $j=1,...,M$. We can then for eact time $t=1,...,T$ define the $N\times M$ matrix $D_t$ as such

$$
D_t=\begin{bmatrix} S_t^1(\omega_1)&\cdots &S_t^1(\omega_M)\\
\vdots &\ddots & \vdots\\
S_t^N(\omega_1) &\cdots&S_t^M(\omega_M)\end{bmatrix}.
$$

We will assume that $S_0^1>0$ and $S_1^1(\omega_j)>0$, $j=1,...,M$.

##### Absence of Arbitrage

We now define a **portfolio** as an $N$-dimensional row vector

$$
h=\begin{bmatrix} h^1, \dots,h^N\end{bmatrix}
$$

representing the amount of assets held at time $t=0$ and held until $t=1$. The **value process** is then

$$
V^h_t=h\cdot S_t=\sum_{i=1}^N h^iS_t^i,\ t=0,1.
$$

For a given $\omega_j\in\Omega$ we have the realisation

$$
V_t^h=hS_t(\omega_j)=hd_j=(hD)_j.
$$

<blockquote class = "def">

**Definition 3.1.** The portfolio $h$ is an **arbitrage portfolio** fil it satisfies the conditions: $V_0^h=0$, $P(V_1^h\ge 0)=1$ and $P(V_1^h>0)>0$.

</blockquote>

<blockquote class = "lem">

**Lemma 3.2.** **(Farkas' Lemma)** Suppose that $d_0,d_1,...,d_M$ are column vectors in $\mathbb{R}^N$. Then exactly one of the following problems possesses a solution.

  * **Problem 1**: There exist $\lambda_1,...,\lambda_M\ge0$ such that $d_0=\sum_{j=1}^M\lambda_jd_j$.
  * **Problem 2**: There exist $h\in\mathbb{R}^N$ such that $h^\top d_0<0$ and $h^\top d_j\ge 0$ for $j=1,...,M$.

</blockquote>

We now investegate this system for any possible arbitrage portfolios. However first we acknowledge that there exist a nominal price system $S_t$ and a normalised price system $Z_t$. The latter we define as the nominel pricess under the numeraire $S_t^1$ that is

$$
Z_t=\begin{bmatrix} S_t^1/S_t^1\\
S_t^2/S_t^1\\
\vdots\\
S_t^N/S_t^1\end{bmatrix}=\begin{bmatrix} 1\\
S_t^2/S_t^1\\
\vdots\\
S_t^N/S_t^1\end{bmatrix}.
$$

The reason for introducing the normalized price system is that we can without much effort translate results in this system to the nominal system and the normalised system is easier to analize. For this, however, we need af few results.

<blockquote class = "lem">

**Lemma 3.3.** With notation as above, the following hold.

  1. The $Z_t$ value process i related to the $S_t$ value process by
  $$
  V_t^{h,Z}=hZ_t=\frac{1}{S_t^1}V_t^h.
  $$
  2. A portfolio is an arbitrage in the $S_t$ system if and only if there is an arbitrage in the $Z_t$ system.
  3. In the $Z_t$ price system, the numeraie asset $Z^1$ has unit constant prices i.e. $Z_t^1=1$ for all $t\ge 0$.

</blockquote>

One of the reason that the normalised system is attractable is that the numeraire asset is constant i.e. risk free in the normalised system. Let us formulate our first main result.

<blockquote class = "prop">

**Proposition 3.4.** The market is arbitrage free if and only iff there exists strictly positive real numbers $q_1,...,q_M\ge 0$ with $q_1+\cdots + q_M=1$ (probability vector) such that the following vector equality holds

$$
\begin{bmatrix} Z_0^1\\
\vdots\\
Z_N^1\end{bmatrix}=\begin{bmatrix} Z_1^1(\omega_1)\\
\vdots\\
Z_1^N(\omega_1)\end{bmatrix}q_1+\cdots +\begin{bmatrix} Z_1^1(\omega_M)\\
\vdots\\
Z_1^N(\omega_M)\end{bmatrix}q_M.\tag{3.3}
$$

</blockquote>

<details>
<summary>**Proof.**</summary>

</details>

##### Martingale Measures

<blockquote class = "def">

**Definition 3.5.** Given the objective probability measure $P$ on $(\Omega,\mathcal{F},P)$, we say that another probability measure $Q$ defined on $\Omega$ is  **equivalent** to $P$ if

$$
\forall A\in\mathcal{F}:P(A)=0\iff Q(A)=0,
$$

or equivalently

$$
\forall A\in\mathcal{F}:P(A)=1\iff Q(A)=1.
$$

</blockquote>

<blockquote class = "def">

**Definition 3.7.** Consider the market model above and set $S^1$ as the numeraire asset. We say that a probability measure $Q$ defined on $\Omega$ is a **martingale measure** if it satisfies the following conditions:

  1. $Q$ is equivalent to $P$, i.e. $Q\sim P$.
  2. For every $i=1,...,N$, the normalized asset price process
  $$
  Z_t^i=\frac{S_t^i}{S_t^1},
  $$
  is martingale under the measure $Q$.

</blockquote>

<blockquote class = "thm">

**Theorem 3.8.** **(First Fundamental Theorem)** Given a fixed numeraire, ther market is free of arbitrage possibilities if and only if there exists a martingale measure $Q$.

</blockquote>

By assuming that the numeraire asset is risk free (i.e. does not depend on $\omega$) then by scaling we can derive the short interest rate as

$$
1+R=\frac{S_1^1}{S_0^1}.
$$

With this in mind we can formulate theorem 3.8 in its more widely used form.

<blockquote class = "thm">

**Theorem 3.9.** **(First Fundamental Theorem)** Assume that there exist a risk free asset, and denote the corresponding risk free interest rate by $R$. Then the market is arbitrage free if and only if there exist a measure $Q\sim P$ such that

$$
S_0^i=\frac{1}{1+R}E^Q[S_1^i],\hspace{20pt}\text{for all}\ i=1,...,N.\tag{3.9}
$$

</blockquote>

##### Martingale Pricing

Moving forward we will assume that there exist a risk free asset and we will denote it by $B_t$ ($B_t=S^1_t/S^1_0$).

<blockquote class = "def">

**Definition 3.10.** A **contingent claim** is any random variable $X$, defined on the sample space $\Omega$.

</blockquote>

To ensure no arbitrage in the extended market containing the $N$ assets and the contingent claim we can apply the first fundamental pricing theorem on the extended market.

<blockquote class = "prop">

**Proposition 3.11.** Consider a given claim $X$. In order to avoid arbitrage, $X$ must then be priced according to the formula

$$
\Pi_0[X]=\frac{1}{1+R}E^Q[X],\tag{3.10}
$$

where $Q$ is a martingale measure for the underlying market $(\Pi,S^1,...,S^N)$.

</blockquote>

##### Completeness

Given that a market is arbitrage-free we may run into a uniqueness issue when determining the price of a contingent claim. If a martingale measure exist we will very much like it to be unique as this will ensure that the price from the risk neutral valuation formula is unique. To this we need the market to be complete.

<blockquote class = "def">

**Definition 3.12.** Consider a contingent claim $X$. If there exists a portfolio $h$, based on the underlying assets, such that

$$
V_1^h=X,\ \text{with probability 1}\tag{3.11}
$$

i.e.

$$
V_1^h(\omega_j)=X(\omega_j),\ j=1,...,M,\tag{3.12}
$$

then we say that $X$ is **replicated**, or **hedged** by $h$. Such a portfolio $h$ is called a replicating, or hedging portfolio. If every contingent claim can be replicated, we say that the market is **complete**.

</blockquote>

We can now formulate a proposition on when the market is complete in terms of the matrix $D$.

<blockquote class = "prop">

**Proposition 3.13.** The market is complete if and only if the rows of the matrix $D$ span $\mathbb{R}^M$, i.e. if and only if $D$ has rank $M$.

</blockquote>

Now we formulate the second fundamental pricing theorem in terms of the martingale measure $Q$.

<blockquote class = "prop">

**Proposition 3.14.** **(Second Fundamental Theorem)** Assume that the model is arbitrage free i.e. $Q$ exist. Then the market is unique if and only if the martingale measure is unique.

</blockquote>

##### Stochastic Discount Factors

<blockquote class = "def">

**Definition 3.16.** The random variable $L$ on $\Omega$ is defined by

$$
L(\omega_i)=\frac{q_i}{p_i},\hspace{20pt} i=1,...,M.
$$

</blockquote>

<blockquote class = "def">

**Definition 3.17.** Assume the absence of arbitrage, and fix a martingale measure $Q$. With notation as above, the **stochastic discount factor** (or "state price deflator") is the random variable $\Lambda$ on $\Omega$ by

$$
\mathbf{M}(\omega)=\frac{1}{1+R}\cdot L(\omega).\tag{3.19}
$$

</blockquote>

<blockquote class = "prop">

**Proposition 3.18.** The arbitrage free price of any claim $X$ is given by the formula

$$
\Pi_0[X]=E^P[\mathbf{M}\cdot X]
$$

where $\mathbf{M}$ is a stochastic discount factor.

</blockquote>

&nbsp;

### Exercises

**Probability exercises**

Let $(W(t))_{t\ge}$ be a Brownian motion (Bjork, Definition 4.1).

**Exercise 1.** Show that the following processes also are Brownian motions.

 i. $(-W(t))_{t\ge 0}$ (symmetry)
 ii. For any $s\ge 0$, $(W(t+s)-W(s))_{t\ge 0}$ (time-homogeneity).
 iii. For every $c>0$, $(cW(t/c^2))_{t\ge 0}$ (scaling).

<details>
<summary>**Solution (i).**</summary>

By assumption $W$ is a Brownian motion and so it follows that

$$-W_0=-1\cdot0=0$$

Furthermore, for $r<s\le t< u$ it holds that $W_u-W_t$ and $W_s-W_r$ is independent. By seperate transformations the independence property is preserved and $-(W_u-W_t)$ and $-(W_s-W_r)$ is independent. Next, for a normal distributed random variable $N\sim\mathcal{N}(\mu,\sigma^2)$ it holds, that for a scaler $c\in\mathbb{R}$ we have $c N\sim\mathcal{N}(c\mu,c^2\sigma ^2)$. Then obviously;

$$-(W_t)=(-1)W_t\stackrel{d}{=}\mathcal{N}((-1)\cdot0,(-1)^2(t-s))\stackrel{d}{=}\mathcal{N}( 0,t-s).$$

Lastly, let $\omega \in \Omega$ and consider the sample path $s\mapsto (-W_s)(\omega)$. Clearly for two continuous functions $f$ and $g$ it holds that $(g\circ f)$ is continuous. Then with $g(f)=-f$ and $f(t)=W_t(\omega)"/>$ it follows that $(-W_t)=(g\circ W)(t)$ is also continuous.

</details>
<details>
<summary>**Solution (ii).**</summary>

Much like the previous exercise we define a new process and show the properties hold. Let $s\ge 0$ be chosen arbitrary. Now define $X_t=W(t+s)-W(s)$.

First, we let $t=0$ and see

$$X_0=W(0+s)-W(s)=W(s)-W(s)=0.$$

Secondly, we have that for $r<u$:

$$X_u-X_r=W(u+s)-W(s)-(W(r+s)-W(s))=W(u+s)-W(r+s)\sim \mathcal{N}(0,u+s-(r+s))=\mathcal{N}(0,u-r).$$

and since for $r<u\le k<l$ the translation $r+s<u+s\le k+s<l+s$ still holds and $X_l-X_k=W(l+s)-W(k+s)$ and $X_u-X_r=W(u+s)-W(k+s)$ are independent. Finally since $W_t(\omega)$ is continuous in $t$ hence the translation $W_{t+s}$ is continouos. Adding a constant yields a function that is also continuous, hence $X_t$ is continuous.

</details>
<details>
<summary>**Solution (iii).**</summary>

Let $c>0$ be given. We show that

$$X_t=cW\left(\frac{t}{c^2}\right)$$

is a Brownian motion. We simply show the four properties. Let $t=0$ and notice

$$X_0=cW\left(\frac{0}{c^2}\right)=cW(0)=0.$$

The second property follows from seperate transformation and that for $r<u\le s<t$ we consider

$$X_u-X_r=c\left(W\left(\frac{u}{c^2}\right)-W\left(\frac{r}{c^2}\right)\right)\hspace{20pt}\text{and}\hspace{20pt}X_t-X_s=c\left(W\left(\frac{t}{c^2}\right)-W\left(\frac{s}{c^2}\right)\right)$$

and since $c,r,u,t,s>0$ we have the same order for the scaled version of $r,u,t,s$ and hence we have two independent RV scaled by $c$. Then by seperate transformations the variables is still independent. Next for the third property:

$$X_t-X_s=c\left(W\left(\frac{t}{c^2}\right)-W\left(\frac{s}{c^2}\right)\right)\sim\mathcal{N}\left(c\cdot 0,c^2\left(\frac{t}{c^2}-\frac{s}{c^2}\right)\right)=\mathcal{N}(0,t-s).$$

Where we use the properties of scaling a normal distributed random variable i.e. for $c>0$ and $N\sim\mathcal{N}(\mu,\sigma ^2)$ it follows that $c N\sim\mathcal{N}(c\mu,c^2\sigma ^2)$. Finally, the forth property follows since $g(f)=cf$ is continuous and $h(t)=t/c^2$ is continuous, then for any continuous function $f(s)$ it follows that $(g \circ f\circ h)=g(f(h(t)))$ is continuous.

</details>

&nbsp;

<blockquote class = "prop">
**Proposition B.37.** Let $(\Omega,\mathcal{F},P)$ be a given probability space, let $\mathcal{G}$ be a sub-sigma-algebra of $\mathcal{F}$, and let $X$ be a square integrable random variable.
Consider the problem of minimizing
$$E\left[(X-Z)^2\right]$$
where $Z$ is allowed to vary over the class of all square integrable $\mathcal{G}$ measurable random variables. The optimal solution $\hat{Z}$ is then given by.
$$\hat{Z}=E[X\vert\mathcal{G}].$$
</blockquote>

**Exercise 2.** *(Bjork, exercise B.11.)* Prove proposition B.37 by going along the following lines.

  a. Prove that the "estimation error" $X-E[X\vert\mathcal{G}]$ is orthogonal to $L^2(\Omega,\mathcal{G},P)$ in the sence that for any $Z\in L^2(\Omega,\mathcal{G},P)$ we have
  $$E[Z\cdot(X-E[X\vert\mathcal{G}])]=0$$
  b. Now prove the proposition by writing
  $$X-Z=(X-E[X\vert\mathcal{G}])+(E[X\vert\mathcal{G}]-Z)$$
  and use the result just proved.

<details>
<summary>**Solution (a).**</summary>

Let $X\in L^2(\Omega,\mathcal{F},P)$ be a random variable. Now consider an arbitrary $Z\in L^2(\Omega,\mathcal{G},P)$. Recall that $\mathcal{G}\subset \mathcal{F}$ and so $X$ is also in $Z\in L^2(\Omega,\mathcal{G},P)$, as it is bothe square integrable and $\mathcal{G}$-measurable. Then

$$E\left[Z\cdot(X-E[X\vert\mathcal{G}])\right]=E\left[Z\cdot X\right]-E\left[Z\cdot E[X\vert\mathcal{G}]\right].$$

Then by using the law of total expectation and secondly that $Z$ is $\mathcal{G}$-measurable we have that

$$E\left[Z\cdot X\right]=E\left[E[Z\cdot X\vert\mathcal{G}]\right]=E\left[Z\cdot E[ X\vert\mathcal{G}]\right].$$

Combining the two equations gives the desired result.

</details>
<details>
<summary>**Solution (b).**</summary>

Obviously, we have that

$$X-Z=X-Z+E[X\vert\mathcal{G}]-E[X\vert\mathcal{G}]=(X-E[X\vert\mathcal{G}])+(E[X\vert\mathcal{G}]-Z).$$

Then squaring the terms gives

$$(X-Z)^2=(X-E[X\vert\mathcal{G}])^2+(E[X\vert\mathcal{G}]-Z)^2+2(X-E[X\vert\mathcal{G}])(E[X\vert\mathcal{G}]-Z)$$

Taking expectation on each side and using linearity of the expectation we have that

$$E[(X-Z)^2]=E\left[(X-E[X\vert\mathcal{G}])^2\right]+E\left[(E[X\vert\mathcal{G}]-Z)^2\right]+2E\left[(X-E[X\vert\mathcal{G}])(E[X\vert\mathcal{G}]-Z)\right].$$

We can now use that $E[X\vert\mathcal{G}]-Z$ is $\mathcal{G}$-measurable with the above result on the last term.

$$E[(X-Z)^2]=E\left[(X-E[X\vert\mathcal{G}])^2\right]+E\left[(E[X\vert\mathcal{G}]-Z)^2\right].$$

Now since $X$ is given the term $E\left[(X-E[X\vert\mathcal{G}])^2\right]$ is simply a constant not depending on the choice og $Z$. The optimal choice of $Z$ is then $E[X\vert\mathcal{G}]$ since this minimizes the second term. The statement is then proved.

</details>

&nbsp;

**Exercise 3.** Discuss the following theory/results of Moment generating functions (Laplace transform).

Let $X$ be a random variable with distribution function $F(x)=P(X\le x)$ and $Y$ be a random variable with distribution function $G(y)=P(Y\le y)$.

<blockquote class = "def">
**Definition.** The moment generating function or Laplace transform of $X$ is

$$\psi_X(\lambda)=E\left[e^{\lambda X}\right]=\int_{-\infty}^\infty e^{\lambda x}dF(x)$$

provided the expectation is finite for $\vert\lambda\vert<h$ for some $h>0$.
</blockquote>

The MGF uniquely determine the distribution of a random variable, due to the following result.

<blockquote class = "thm">
**Theorem 1.** *(Uniqueness)* If $\psi_X(\lambda)=\psi_Y(\lambda)$ when $\vert\lambda\vert<h$ for some $h>0$, then $X$ and $Y$ has the same distribution, that is, $F=G$.
</blockquote>

There is also the following result of independence for Moment generating functions.

<blockquote class = "thm">
**Theorem 1.** *(Independence)* If 

$$E\left[e^{\lambda_1X+\lambda_2Y}\right]=\psi_X(\lambda_1)\psi_Y(\lambda_2)$$

for $\vert\lambda_i\vert<h$ for $i=1,2$ for some $h>0$, then $X$ and $Y$ are independent random variables.
</blockquote>

**Example.** Recall that the Moment generating function of a normal (Gaussian) distribution is given by

$$\psi_X(\lambda)=E\left[e^{\lambda X}\right]=\exp\left(\lambda \mu + \frac{\lambda^2}{2}\sigma^2\right)$$

where $X$ is normally distributed with mean $\mu$ and variance $\sigma^2$ and $\lambda\in\mathbb{R}$ is a constant. Since a Brownian motion $W(t)$ is normally distributed with zero mean and variance $t$, we have that

$$E[\exp(\lambda W(t))]=\exp\left(\frac{\lambda^2}{2}t\right).$$

<details>
<summary>**Discussion.**</summary>



</details>

&nbsp;

**Exercise 4.** *(Bjork, exercise C.8.(a-c))* Let $W$ be a Brownian motion. Notice that for the natural filtration $\mathcal{F}_s=\sigma(W_t\vert t\le s)$ $W_t-W_s$ is independent of $\mathcal{F}_s$

  a. Show that $W_t$ is a martingale.
  b. Show that $W^2_t-t$ is a martingale.
  c. Show that $\exp(\lambda W_t-\frac{\lambda^2}{2}t)$ is a martingale.

<details>
<summary>**Solution (a).**</summary>

We show that for the natural filtration that $W_t$ is a martingale. This include showing integrability and the martingale property. For the first we note that for a normal distributed random variable with mean 0 we have

$$E[\vert N\vert]=\int_{-\infty}^\infty \vert x\vert dF_N(x)=2\int_{0}^\infty xdF_N(x)$$

since the distribution is symmetric. Substituting the distribution function $\Phi(x)=P(N\le x)$ in we see that

$$E[\vert N\vert]=2\int_{0}^\infty xd\Phi(x)=2\int_{0}^\infty x\frac{1}{\sqrt{2\pi\sigma^2}}e^{-x^2/(2\sigma^2)}dx=(*)$$

by substituting $u=x^2/(2\sigma^2)$ ($x=\sqrt{2\sigma^2u}$) we have that

$$\frac{dx}{du}=\frac{1}{2}\sqrt{2\sigma^2u}2\sigma^2=(\sigma^2)^{3/2}\sqrt{2}u\iff dx=(\sigma^2)^{3/2}\sqrt{2}u\ du$$

hence

$$(*)=\frac{2}{\sqrt{2\pi\sigma^2}}\int_0^{\infty}\sqrt{2\sigma^2u}e^{-u}(\sigma^2)^{3/2}\sqrt{2}u\ du=\frac{2\sqrt{2\sigma^2}(\sigma^2)^{3/2}\sqrt{2}}{\sqrt{2\pi\sigma^2}}\int_0^{\infty}\sqrt{u}e^{-u}u\ du.$$

This then simplify to

$$(*)=\frac{(2\sigma^2)^{3/2}}{\sqrt{\pi}}\int_0^{\infty}u^{3/2}e^{-u}\ du=(2\sigma^2)^{1/2}\sqrt{\frac{2\sigma^2}{\pi}}\int_0^{\infty}u^{3/2}e^{-u}\ du=\sqrt{\frac{2\sigma^2}{\pi}}<\infty.$$

(Obviously the above is not derived correctly, but the end expression is valid, source: [link](https://arxiv.org/pdf/1402.3559.pdf)) However since

$$W_t=W_t-0=W_t-W_0\sim\mathcal{N}(0,t)$$

we have that $E\vert W_t\vert<\infty$ as desired.

Next, we have that

$$E[W_t\vert \mathcal{F}_s]=E[W_t-W_s\vert\mathcal{F}_s]+W_s=0+W_s=W_s.$$

In the above we used that $W_t-W_s$ is $\mathcal{F}_s$-measurable with mean 0. Then it follows that $W_t$ is a martingale.

</details>
<details>
<summary>**Solution (b).**</summary>

Let $M_t=W_t^2-t$. First, we observe that two measurable functions composed is still a measurable function. Hence we know that $M_t$ is measurable wrt. the filtration since $W_t$ is measurable and $w\mapsto w^2+t$ is measurable. Secondly, we have that

$$E[\vert W_t^2-t\vert]\le E\vert W_t^2\vert +E\vert t\vert=t+t=2t<\infty$$

where we use the triangle inequality. Thirdly, for the martingale property we have that for $t>s$:

$$E[M_t\vert \mathcal{F}_s]=E[W_t^2-t\vert \mathcal{F}_s]=E[W_t^2+W_s^2-2W_tW_s-W_s^2+2W_tW_s-t\vert \mathcal{F}_s]$$

which by linearity and independence of increments to the filtration gives

$$E[M_t\vert \mathcal{F}_s]=E[(W_t-W_s)^2-W_s^2+2W_tW_s-t\vert \mathcal{F}_s]=t-s-t+E[2W_tW_s-W_s^2\vert \mathcal{F}_s]$$

However since $W_s$ is measurable wrt. the filtration at time $s$ the above is

$$E[M_t\vert \mathcal{F}_s]=2W_sE[W_t\vert \mathcal{F}_s]-W_s^2-s=2W_s^2-W_s^2-s=W_s^2-s=M_s.$$

Since from (a) we know that $W_t$ is a martingale. Then we arrive at the desired result.

</details>
<details>
<summary>**Solution (c).**</summary>

Let $M_t=\exp\left(\lambda W_t-\frac{\lambda^2}{2}t\right)$. First, by composition of measurable functions $M_t$ is $\mathcal{F}_t$-measurable. Secondly, we have using the MGF for a normal distributed random variable:

$$E\vert M_t=E\left(\exp\left(\lambda W_t-\frac{\lambda^2}{2}t\right)\right)\le E\left(\exp\left(\lambda W_t\right)\right)=\exp\left(\frac{\lambda^2}{2}t\right)<\infty.$$

Thirdly, we consider

$$E[M_t\vert\mathcal{F}_s]=E\left.\left[\left(\exp\left(\lambda W_t-\frac{\lambda^2}{2}t\right)\right)\right\vert\mathcal{F}_s\right]=\exp\left(-\frac{\lambda^2}{2}t\right)E\left.\left[\left(\exp\left(\lambda W_t\right)\right)\right\vert\mathcal{F}_s\right].$$

By adding and subtracting $W_s$ in the exponent we get

\begin{align*}
E[M_t\vert\mathcal{F}_s]&=\exp\left(-\frac{\lambda^2}{2}t\right)E\left.\left[\left(\exp\left(\lambda (W_t-W_s)+\lambda W_s\right)\right)\right\vert\mathcal{F}_s\right]\\
&=\exp\left(-\frac{\lambda^2}{2}t\right)\exp\left(\frac{\lambda^2}{2}(t-s)\right)E\left.\left[\left(\exp\left(\lambda W_s\right)\right)\right\vert\mathcal{F}_s\right].
\end{align*}

Using that $E\left.\left[\left(\exp\left(\lambda W_s\right)\right)\right\vert\mathcal{F}_s\right]=\exp\left(\lambda W_s\right)$ and combining the exponents gives the desired:

$$E[M_t\vert\mathcal{F}_s]=\exp\left(\lambda W_s-\frac{\lambda^2}{2}s\right)=M_s.$$

</details>

## Week 2 

### Material

  * Stochastic integrals and Ito formula (Chapter 4 and Appendix C.2)
  * Stochastic differential equations (Chapter 5.1-4)

### Theory

#### Stochastic Integrals

#### Discrete Stochastic Integrals

#### Stochastic Differential Equations

### Exercises

**Exercise 4.1** *(Bjork)* 

## Week 3

### Material

  * Partial differential equations (Chapter 5.5)
  * Self-financing portfolios (Chapter 6)
  * Black-Scholes PDE (classic approach) and risk neutral valuation (Chapter 7.1-5)

### Theory

### Exercises

## Week 4

### Material

  * Black-Scholes formula (Chapter 7.6, see also Remark to Black-Scholes formula)
  * Completeness and hedging (Chapter 8)
  * Put-call parity (Chapter 10.1)
  * The Greeks (Chapter 10.2)
  * Risk neutral valuation formula (Chapter 11.6) 
  * Equivalent probability measures (Appendix A.11, B.6 and C.3) 

### Theory

### Exercises

## Week 5

### Material

  * Girsanov theorem (Chapter 12, see also Levy characterization of Brownian motion and proof of Girsanov)
  * Martingale representation theorem (Chapter 12)

### Theory

### Exercises

## Week 6

### Material

  * Black-Scholes model, martingale approach (Chapter 13)
  * Multidimensional models (Chapter 14)

### Theory

### Exercises

## Week 7

### Material

  * Pricing and proof of fundamental pricing theorem I and II (Chapter 11)
  * Incomplete Markets (Chapter 9)

### Theory

### Exercises
