---
title: "Homework (FinKont)"
output:
  html_document:
    toc: false
    toc_float: true
    code_folding: hide
---

```{r, include=FALSE, results = 'hide'}
library(ggplot2)
library(dplyr)
```


# Introduction

# Weeks {.tabset}

## Week 1

### Material

  * Brownian motion (Chapter 4.1)
  * Conditional expectation (Appendix B.5) 
  * Filtration (Appendix B.3 and Chapter 4.2)
  * Martingales (Appendix C.1 and Chapter 4.4)
  * Introduction (Chapter 1)
  * Discrete time models (Chapter 2 and 3)

### Theory

>**Definition 4.1.** *(Brownian motion)* A stochastic process $W$ is called a **Brownian motion** or **Wiener process** if the following conditions hold
>
> 1. $W_0=0$.
> 2. The process $W$ has independent increments, i.e. if $r<s\le t< u$ then $W_u-W_t$ and $W_s-W_r$ are independent random variables.
> 3. For $s<t$ the random variable $W_t-W_s$ has the Gaussian distribution $\mathcal{N}(0,t-s)$.
> 4. $W$ has continuous trajectories i.e. $s\mapsto W(s;\omega)$ i continuous for all $\omega \in\Omega$.

```{r}
#Example of trajectory for BM
set.seed(1)
t <- 0:1000
N <- rnorm(
  n = length(t)-1, #initial value = 0
  mean = 0, #incements mean = 0
  sd = sqrt(t[2:length(t)] - t[1:(length(t)-1)]) #increment sd = sqrt(t-s)
)
W <- c(0,cumsum(N))
```
```{r,echo=FALSE,out.width= "50%", out.extra='style="float:right; padding:10px"',include=TRUE}
data.frame(t = t, W = W) %>%
  ggplot() +
  geom_line(aes(x=t,y=W)) +
  labs(title = "Realisation of a Brownian motion") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        title = element_text(size = 18))
```

### Exercises

**Probability exercises**

Let $(W(t))_{t\ge}$ be a Brownian motion (Björk, Definition 4.1).

**Exercise 1.** Show that the following processes also are Brownian motions.

 i. $(-W(t))_{t\ge 0}$ (symmetry)
 ii. For any $s\ge 0$, $(W(t+s)-W(s))_{t\ge 0}$ (time-homogeneity).
 iii. For every $c>0$, $(cW(t/c^2))_{t\ge 0}$ (scaling).

<details>
<summary>**Solutions (i).**</summary>

By assumption $W$ is a Brownian motion and so it follows that

$$
-W_0=-1\cdot 0=0.
$$

Furthermore, for $r<s\le t< u$ it holds that $W_u-W_t$ and $W_s-W_r$ is independent. By transformation invariance of the independence property $-(W_u-W_t)$ and $-(W_s-W_r)$ is independent.

</details>
<details>
<summary>**Solutions (ii).**</summary>



</details>
<details>
<summary>**Solutions (iii).**</summary>



</details>

## Week 2

### Material

MFE refers to the book by McNeil, Frey, and Embrechts, while HL refers to the notes by Hult and Lindskog posted on Absalon.

 * Var-Cov method, simulation, importance sampling and bootstrap. For the Var-Cov method, see MFE Sec. 9.2. For importance sampling and bootstrap, see the supplementary reading in Absalon.
 * Extreme value theory: MFE Ch. 5 or alternative reading in HL (suggested for this part).

### Theory

### Exercises

## Week 3

### Material

MFE refers to the book by McNeil, Frey, and Embrechts, while HL refers to the notes by Hult and Lindskog posted on Absalon.

 * Spherical and elliptical distributions: MFE Sec. 6.3.
 * Spherical and elliptical distributions, cont.; introduction to copulas.

### Theory

### Exercises

## Week 4

### Material

MFE refers to the book by McNeil, Frey, and Embrechts, while HL refers to the notes by Hult and Lindskog posted on Absalon.

 * Copulas: MFE Sec. 7.1-7.5 (Sklar's theorem, Frechet bounds).
 * Copulas cont. (Transformations; examples; Archimedean copulas).

### Theory

### Exercises

## Week 5

### Material

MFE refers to the book by McNeil, Frey, and Embrechts, while HL refers to the notes by Hult and Lindskog posted on Absalon.

 * Copulas cont. (simulating Archimedean copulas, statistical methods, measures of dependence).
 * Credit risk: the Merton model. MFE Ch. 10, particularly Sec. 10.3. [This lecture will only last two hours.]

### Theory

### Exercises

## Week 6

### Material

MFE refers to the book by McNeil, Frey, and Embrechts, while HL refers to the notes by Hult and Lindskog posted on Absalon.

 * Portfolio credit risk: MFE Ch. 11, specifically Sections 11.1-11.3.
 * Portfolio credit risk, cont. [This lecture will only last two hours.]

### Theory

### Exercises

## Week 7

### Material

MFE refers to the book by McNeil, Frey, and Embrechts, while HL refers to the notes by Hult and Lindskog posted on Absalon.

 * Intro. to operational risk. Stochastic processes in risk management: stochastic models for operational risk; financial time series models. Connections to non-life insurance models and estimates (last lecture).

### Theory

### Exercises
