---
title: "Homework (FinKont)"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
---

```{r, include=FALSE, results = 'hide'}
library(ggplot2)
library(dplyr)
#rmarkdown::render(input = "FinKont_homework.Rmd", output_format = "pdf_document")
```

## Stochastic processes

### Brownian Motion

>**Brownian motion** *(Bjork, def. 4.1)*  A stochastic process $W$ is called a **Brownian motion** or **Wiener process** if the following conditions hold
>
> 1. <img class="math inline"  src="https://math.vercel.app?inline=W_0=0"/>.
> 2. The process <img class="math inline"  src="https://math.vercel.app?inline=W"/> has independent increments, i.e. if <img class="math inline"  src="https://math.vercel.app?inline=r<s\le t< u"/> then <img class="math inline" src="https://math.vercel.app?inline=W_u-W_t"/> and <img class="math inline"  src="https://math.vercel.app?inline=W_s-W_r"/> are independent random variables.
> 3. For <img class="math inline"  src="https://math.vercel.app?inline=s<t"/> the random variable <img class="math inline"  src="https://math.vercel.app?inline=W_t-W_s"/> has the Gaussian distribution <img class="math inline"  src="https://math.vercel.app?inline=\mathcal{N}(0,t-s)" />.
> 4. <img class="math inline"  src="https://math.vercel.app?inline=W"/> has continuous trajectories i.e. <img class="math inline"  src="https://math.vercel.app?inline=s\mapsto W(s;\omega)"/> i continuous for all <img class="math inline"  src="https://math.vercel.app?inline=\omega \in\Omega"/>.

```{r}
#Example of trajectory for BM
set.seed(1)
t <- 0:1000
N <- rnorm(
  n = length(t)-1, #initial value = 0
  mean = 0, #incements mean = 0
  sd = sqrt(t[2:length(t)] - t[1:(length(t)-1)]) #increment sd = sqrt(t-s)
)
W <- c(0,cumsum(N))
```
```{r,echo=FALSE,out.width= "50%", out.extra='style="float:right; padding:10px"',include=TRUE}
data.frame(t = t, W = W) %>%
  ggplot() +
  geom_line(aes(x=t,y=W)) +
  labs(title = "Realisation of a Brownian motion") +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        title = element_text(size = 18))
```
